---
title: "Mental health service utilization during COVID-19 pandemic"
author: "Camille Parchment"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

This is my PM566 Final Project website. file:///C:/Users/camil/Desktop/GitHub/PM-566-final/README.html

<br>

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Initialize code chunk options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
```

# Introduction
  
During the COVID-19 pandemic, mental health services were disrupted, causing profound consequences for individuals in need of consistent mental health care. The conditions of the pandemic i.e., stay at home orders, widespread layoffs, a decrease of mental health providers, and changes to healthcare delivery to comply with safety protocols, may have decreased access to necessary mental health services during the COVID-19 pandemic.

Therefore, the primary aim of this study is to establish insurance-related patterns of mental healthcare utilization during October 28, 2021 – November 9, 2021. Specifically, did the type of health insurance affect access to mental health services during the COVID-19 pandemic?

# Data 

The Household Pulse Survey (HPS) is a self-report online survey which measures in real time the social and economic impacts of the COVID-19 pandemic. In this study we analyzed HPS data from week 18,  October 28, 2021 – November 9, 2021. 
Here is a link to the data https://www.census.gov/programs-surveys/household-pulse-survey/datasets.html#phase3

The primary question to establish receipt of mental health services is "at any time in the last 4 weeks, did you need counseling or therapy from a mental health professional, but DID NOT GET IT for any reason?"

Mental health services are defined as counseling or therapy.

# Study Sample

Our study sample included all individuals who requested mental health services and either received services or did not between October 28, 2021, and November 9, 2021. Individuals were categorized as having either public or private insurance. 

```{r echo=FALSE}
source("mental_health_data.R")

private.insurance <- DF4 %>% 
                     plot_ly(x = ~Total_sv_needed, y = ~num_notrec, 
                     type = 'scatter',
                     mode = 'markers',
                     color = ~state,
                     colors = "Purples")
                  
private.insurance <- private.insurance %>% layout(title = 'Mental health services requested and not recieved by state and private insurance ',
                                                xaxis = list(title = 'Total services requested'),
                                                yaxis = list(title = 'Number of services not recived'))


public.insurance <- DF2 %>% 
                     plot_ly(x = ~Total_sv_needed, y = ~num_notrec, 
                     type = 'scatter',
                     mode = 'markers',
                     color = ~state,
                     colors = "Reds")
                     
public.insurance <- public.insurance %>% layout(title = 'Mental health services requested and not recieved by state and public insurance',
                                                xaxis = list(title = 'Total services requested'),
                                                yaxis = list(title = 'Number of services not recived'))

 income.data$hover_pri <- with(income.data, paste( income, '<br>', income.data$Notrec_rate.x, "rate"))
income.data$hover_pub <- with(income.data, paste( income, '<br>', income.data$Notrec_rate.y, "rate"))

income <- income.data %>%plot_ly
income <- income %>% add_trace(x = ~income.data$income, y = ~income.data$num_notrec.x, 
                                type = 'bar',
                                name = 'Private Insurance',
                                text = income.data$hover_pri,
                                color = 'Blue')
income <- income %>% add_trace( x= ~income.data$income, y= ~income.data$num_notrec.y, 
                                type = 'bar',
                                name = "Public Insurance",
                                text = income.data$hover_pub,
                                color = 'Purple')

income <- income %>% layout( title = 'Mental health services requested and not recived by income and insurance type',
                             yaxis = list( title = 'Number of services requested <br> and not received'), barmode = 'group',
                             xaxis = list (title = "Income"),
                             color = ~income.data$Notrec_rate.x,
                             color = ~income.data$Notrec_rate.y,
                            colors = "Blues")

map.data$hover <- with(map.data, paste(" total services requested", '<br>', state, '<br>', tot.svs.notrec, "total services not received", '<br>',  tot.notrec.rate, " % services not received "  ))



l <- list(color = toRGB("white"), width=2)


g <- list(
     scope = 'usa',
     projection = list(type = 'albers usa'),
     showlakes = TRUE, 
     lakecolor = toRGB('white')
)




fig <- plot_geo(state_data.comp, 
               locationmode = 'USA-states')



fig <- fig %>% add_trace (z = ~map.data$total.needed,    
           text = ~map.data$hover,
       locations = ~map.data$usa_state_code,
           color = ~map.data$Notrec_rate,
          colors = 'Purples' )

              

fig <- fig %>% colorbar(title = 'Rate of services not received')
fig <- fig  %>% layout (title = "Rate of total mental health services requested and not received <br> by state and both public and private insurance type",
         geo =g )
         


```



## Results {.tabset}

The following plots depict the rates and counts of mental health services that were requested and not received by state, insurance type and income.

### Services not received <br> by income
```{r echo= FALSE}
income 
```

This bar graph illustrates the count and rate of mental health services requested and not received in eight income categories. Among those with public insurance, individuals making less than 25k a year have the highest rate of not receiving services (19.7 %) compared to other income levels. Although individuals with private insurance who make between 100k and 150k a year have the highest count of services requested and not received, individual making less than 25k and up to 34.9k have a higher rate of not receiving services (18.92%, 18.91% respectively).

Individuals who have public insurance have lower rates of not receiving services than those who have private insurance with the exception of individuals who make less than 25k a year and those who make between 35k and 49k a year. 

### Private insurance <br> by state

```{r echo=FALSE}
private.insurance
```

This scatter plot illustrates the number of services that were requested and not received during October 28, 2021 – November 9, 2021. It can be seen that California has the highest number of mental health services that were requested and not received during the pandemic  amoung those with private insurance. 

### Public insurance <br> by state 

```{r echo=FALSE}
public.insurance
```

This scatter plot illustrates the number of services that were requested and not received during October 28, 2021 – November 9, 2021. It can be seen that California has the highest number of mental health services that were requested and not received during the pandemic  amoung those with public insurance. 

### Total service <br> not received

```{r echo=FALSE}
fig
```

This map depicts the total rate of mental health services requested and not received for individuals with both public and private insurance by state. Although California had the highest count of services not received among private and public insurance respectively, California has a lower rate of services not received than Utah (14.5), Oregon (15.8), Oklahoma (16.2),and  Vermont (14.6).

## {-}

# Conclusion 

These data indicate that individuals with public insurance have lower rates of not receiving services that those who have private insurance. However, individuals who make less than 25k a year have similar rates of not receiving services for both private and public insurance. Further research is needed to identify service and diagnosis-specific patterns in mental health service utilization. 
