---
title: "Mental health service utilization during COVID-19 Pandemic"
author: "Camille Parchment"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

This is my PM566 Final Project website.

<br>

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Initialize code chunk options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
```

# Introduction
  
During the COVID-19 pandemic, mental health services were disrupted, causing profound consequences for individuals in need of consistent mental health care. The pandemic exacerbated the need for mental health services which expanded an existing unmet demand. In addition to a shortage of mental health professionals, mental health providers needed to adjust service delivery to comply with safety protocols, which further inhibited access to care for some individuals. Further, at the onset of the pandemic, several state and local governments issued stay-at-home orders to reduce viral spread. 

Employers are the leading providers of health insurance in the United States, suggesting that newly unemployed individuals may have lost access to secure healthcare coverage. The growing demand for mental health services, possible reduction in coverage, and changing delivery systems, may have decreased access to necessary mental health services during the COVID-19 pandemic.

Therefore, the primary aim of this study is to establish insurance-related patterns of mental healthcare utilization. Specifically, did the absence of health insurance decrease access to mental health services during the COVID-19 pandemic?

# Data 

The Household Pulse Survey (HPS) is a self-report online survey which measures in real time the social and economic impacts of the COVID-19 pandemic. In this study we analyzed HPS data from week 18,  October 28, 2021 – November 9, 2021. 

# Study Sample

Our study sample included all individuals who requested mental health services and either received services or did not between October 28, 2021, and November 9, 2021. Individuals were categorized as having either public or private insurance. 

```{r echo=FALSE}
source("mental_health_data.R")

private.insurance <- DF4 %>% 
                     plot_ly(x = ~Total_sv_needed, y = ~num_notrec, 
                     type = 'scatter',
                     mode = 'markers',
                     color = ~state,
                     colors = "Purples")
private.insurance <- private.insurance %>% layout(title = 'Mental health services requested and not recieved by state and private insurance ',
                                                xaxis = list(title = 'Total services requested'),
                                                yaxis = list(title = 'Number of services not recived'))


public.insurance <- DF2 %>% 
                     plot_ly(x = ~Total_sv_needed, y = ~num_notrec, 
                     type = 'scatter',
                     mode = 'markers',
                     color = ~state,
                     colors = "Reds")
public.insurance <- public.insurance %>% layout(title = 'Mental health services requested and not recieved by state and public insurance',
                                                xaxis = list(title = 'Total services requested'),
                                                yaxis = list(title = 'Number of services not recived'))

 income <- plot_ly(x = ~income.data$income, y = ~income.data$num_notrec.x, 
                   type = 'bar',
                   name = 'Private Insurance')
income <- income %>% add_trace( y= ~income.data$num_notrec.y, name = "Public Insurance" )
income <- income %>% layout( title = 'Mental health services requested and not recived by income and insurance type',
                             yaxis = list( title = 'Number of services requested and not received'), barmode = 'group',
                             xaxis = list (title = "Income"),
                             color = ~income,
                            colors = "Blues")

map.data$hover <- with(map.data, paste(state, '<br>', tot.notrec.rate, " % services not received " ))



l <- list(color = toRGB("white"), width=2)


g <- list(
     scope = 'usa',
     projection = list(type = 'albers usa'),
     showlakes = TRUE, 
     lakecolor = toRGB('white')
)




fig <- plot_geo(state_data.comp, 
               locationmode = 'USA-states')



fig <- fig %>% add_trace (z = ~map.data$total.needed,    
           text = ~map.data$hover,
       locations = ~map.data$usa_state_code,
           color = ~map.data$Notrec_rate,
          colors = 'Purples' )

              

fig <- fig %>% colorbar(title = 'Rate of services not received')
fig <- fig  %>% layout (title = "Rate of total mental health services requested and not received <br> by state and both public and private insurance type",
         geo =g )
         


```


## Plots {.tabset}

### Private insurance <br> by state

```{r echo=FALSE}
private.insurance
```

### Public insurance <br> by state 

```{r echo=FALSE}
public.insurance
```

### Services not received <br> by income
```{r echo= FALSE}
income 
```

### Total service <br> not received

```{r echo=FALSE}
fig
```

## {-}
